A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}

### 2023-11-04 paper.tex
Commit riportato del 31 ottobre - Spedito in prima revisione

%=================================================================
% Full title of the paper (Capitalized)
\Title{A sustainable approach to tourist signage on (historical) trails}

% MDPI internal command: Title for citation in the left column
\TitleCitation{A sustainable approach to tourist signage on historical trails}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Firstname Lastname $^{1,\dagger,\ddagger}$\orcidA{}, Firstname Lastname $^{2,\ddagger}$ and Firstname Lastname $^{2,}$*}

@@

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Firstname Lastname, Firstname Lastname and Firstname Lastname}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

% Contact information of the corresponding author
\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}

% Current address and/or shared authorship
\firstnote{Current address: Affiliation 3.} 
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Guiding the visitor to appreciate all the attractions of a region is paramount for building a memorable experience. In the literature, we find many articles exploring the use of sophisticated technologies towards such a goal. The case of wilderness resources exhibits a peculiar aspect, since the presence of human artifacts and the signage may damage the experience, and introduce pollutants as well. In this paper, we consider the case of a natural trail targeted to niche tourism to understand the available options for signage aiming to provide directions, guide the exploration, and warn about dangers. The footprints of the various options are evaluated, and we describe the outcomes of an experimental setup implemented to verify the applicability of our approach.}
%\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results that are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}

@@

%\newcites{Main}{References}
%\newcites{Sitology}{Sitology}

\tableofcontents

@@

Although an official decision is still pending, there is a widely agreed proposal for the naming of the time that we are living as the Anthropocene.  Behind this proposal is the recognition that humans have the potential to significantly alter the environment in which they live on a global scale.

The problem with the above fact is that we humans are not able to understand long-term consequences of deploying such a potential, nor even to keep it under control. Even worse, such capacities are often used with the aim of short-term local effects, ignoring long-term global ones, with the risk of long-term global deterioration. This attitude pervades all activities, starting with the most basic ones such as the supply of food and homes, and is often influenced by economic profit.

@@

To manage this situation, international agencies promote the keyword "sustainability" as a guideline for the conservation of our habitat. Jeffry Ramsey et al. \cite{ram15a} consider the concept behind this word as "vague and disputed but not meaningless" and that cannot be used without a concrete framework, especially in a normative context. To solve to problem of measuring sustainability, Tom Kuhlman et al. \cite{kuh10a} point out an inherent tension in the term between growth and stability. This tension has been attenuated by a change in the meaning of the word, which has gone from focusing on preservation for an indefinite future to a concept that includes three "pillars", society, economy, and environment, as determinants that need to be protected in the present to promote sustainable development in the future. The authors relate artificial and environmental goods, discriminating the two cases where a loss in the latter can or cannot be compensated by an increase in the former, introducing the terms of {\em weak} and {\em strong} sustainability.

The promotion of human cultural heritage is situated in such a complex scenario. As noted by Jorge Otero in \cite{ote22a}, local communities benefit from the promotion of heritage: conservation activity generates social growth, while tourism provides an extra income that can be reinvested in promotion. However, just as in the case of ski resorts, there is a risk of compromising the resource itself. This interaction raises a question of sustainability also for cultural conservation initiatives, two issues that are usually considered in synergy. As in the document "World Heritage and Sustainable Development" of UNESCO, which calls for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".

This research stems from such a painful contradiction in trying to answer the question: is there a way to unlock the potential of a cultural resource while at the same time limiting the damage to a social, economic, and environmental equilibrium?

The question arises in the course of a project funded by the Italian Ministry of Research aimed at applying cutting-edge technologies to the investigation of karst caves that may have been inhabited in the past, from prehistory to today. Among the expected results of the project is the definition of a strategy to disseminate the results of the research and allow non-specialists to visit and enjoy the results, thus contributing to the economy of rural areas that are severely exposed to de-population. A sustainable strategy is even more crucial as the geological environment of interest is particularly fragile, as noted by Aleksander Antic\`c et al. in \cite{ant20a}.

This document provides a conceptual solution for a specific aspect. The solution is conceptual in that it tries to follow a guideline that takes into account as much detail as possible in a way that can be used in other environments but is very practical in its application. Given the closely interlinked nature of the situation, we observe how a very specific aspect, tourist signage, exhibits effects on society, the economy, and the environment.

@@

The functions of signage are twofold
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point to this end, but not all visitors feel comfortable reading a map. The ideal signage should include visible reference points and explicit actions, like "turn right after crossing the stream". A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}). The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail
\item a sign should explain the reasons for interest in a site. This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}). In this sense, the CAI signal above is not useful. A simple board with a site name may be sufficient only when the site is sufficiently popular. Otherwise, a more structured message should be used to explain the reasons why the feature is regarded as relevant.
\end{itemize}

In our study, we consider a range of solutions that help towards sustainable support to sustainable tourism, specifically to the dynamic provision of information during a visit along a natural trail, not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}). We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.

The purpose of such an installation is guiding the {\em user} on a tour that includes urban streets, buildings, natural trails, and caves (this latter being the topic of the project giving financial support to our research). The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as profitable and enjoyable as possible. A traditional approach makes use of physical information boards with graphical or textual content, as in figure \ref{fig:traditional}.

Our study starts by pointing out the issues related to such a solution:

\begin{itemize}
	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports
	\item environmental impact: to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site 
	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
	\item update limits: to update the content, the plaque must be replaced
	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation
\end{itemize}

Such considerations motivate an interest in an alternative way of communication.

@@

We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention in the environment, but the impact of such intervention must be better than that related to traditional signs. To this end, we include in our solution tools that are not part of the site but remain with the user.

During the last decades, we witnessed the diffusion of smartphone devices, and we have no reason to expect a change in such a trend. Smartphones empower individual communication capabilities, allowing them to receive sound, visual, and tactile interactions. The relevance of such capabilities for the improvement of a tourist experience has been widely investigated, either in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a} to find ways to exploit such tools. Again, we start from the limits of such technology, especially in the case of outdoor activities.

@@

Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need to familiarize themselves with new applications.

@@

On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the words of Pearce and Gretzel in \cite{pea12a}, may evoke contrasting feelings, from rewarding to threatening.

@@

The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support, taking into account "by design" its sustainability. Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies the feasibility of a specific solution.

We stress that this article does not aim now at quantifying user satisfaction or the economic revenue associated with the specific solution. Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve further investigation. We aim at isolating an issue, proposing a sustainable strategy for its solution, and implementing a proof of concept for it.

@@

We aim to design sustainable and effective smartphone-based signage.

A straightforward approach concentrates on the wireless capabilities of such a device. Given the premise that the Internet is not reachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest. Small servers connected to the network would provide specific Web services. The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single-chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$. They can provide WiFi Access Points as well as Web content. The ESP8266 has a capacity of 32KB, which is sufficient for explanatory text and a low-resolution image. Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.

Power supply dependency is a serious limitation for such kind of solution. A radio transmitter is a rather power-consuming device, in the range of Watts. Even if intermittent, the operation of a battery-operated networking device cannot be guaranteed for long periods and the host organization should consider power harvesting, which negatively contributes to the economic cost and environmental footprint of the device.

@@

An alternative consists of the utilization of passive devices, like Near Field Communication (NFC) transmitters. The transmitting device is flat, the size of a coin, and costs less than one dollar per piece. To receive the content the smartphone must be very close to the NFC tag. The power needed to operate the radio is drained from the smartphone so that the transmitter does not depend on batteries. The NFC device capacity is in the range of the KBytes, nearly a page on this journal. Storing content in an NFC tag requires a smartphone with a specific app.

The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text which can be treated as such. The smartphone can read it aloud, to compensate user's inabilities, or translated, to cope with linguistic issues. Such capabilities do not need an Internet connection.

@@

\begin{itemize}
	\item {\em Equitable use} is closely related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone
	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation
	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety. The deliberate or accidental release of the passive device into the environment determines a minor pollution
	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
	\item{\em Size and Space for approach and use} need to be carefully considered. In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort
\end{itemize}

@@

As anticipated in the introduction, our study wants to provide a proof of concept for a sustainable solution in a concrete setup. To comply with such a holistic approach, we need to start with the definition of the operation framework.

This section is devoted to the description of an economic and social context, as well as the historical background representing the heritage resource we want to promote. Next, we will discuss the technical details of the solution.

The geographical area of interest is the surroundings of \emph{Casoli}, a small village in a mountainous region in the north of Tuscany, in Italy. The area falls within the municipality of \emph{Bagni di Lucca}, in the province of \emph{Lucca}. In 2021, the archaeological team of our project realized an archaeological map assessing the reasons of interest for the heritage and the geomorphological features of the area.

@@

\emph{Bagni di Lucca} is located on the north-eastern boundary of the Province of \emph{Lucca}, in \emph{Val di Lima}, and is part of the \textit{Media Valle del Serchio} district. With its mountains it marks the historical border with the Modena and Pistoia area, it is very rich in potential for its naturalistic, archaeological, and historical heritage, both expressed and as yet unexpressed. Thanks to the archaeological map the main sites of interest from prehistoric to contemporary times in this municipality are cataloged, photographed, and georeferenced.

From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress to the area a particular geomorphology, very impervious despite the modest hilly elevations.

The \emph{Lima} stream has characterized the history of \emph{Bagni di Lucca} since ever: some of the oldest evidence of human presence in the valley has been found along its ancient river terraces, such as caves and rock shelters frequented since the Palaeolithic age; the manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited since the Middle Age until the 1980s \cite{men76, bed05, ser21}.

On the naturalistic side, the region of \emph{Bagni di Lucca} encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \cite{nat00} European Economic Community (EEC) initiative.

There are three SCIs-SACs (Sites of Community Interest and Special Areas of Conservation) located in the area north of the \emph{Lima} stream, corresponding to the Apennine portion, covering 23\% of the municipal surface:
\begin {itemize}
\item the limestone areas of \emph{Val di Lima} and Balzo Nero; 
\item Monte Prato Fiorito-Monte Coronato-Valle dello
Scesta; 
\item Orrido di Botri.
\end{itemize}

@@

\begin{figure}
\subfloat[Italy and Tuscany, bordered in red]{
	\includegraphics[width = 0.3\linewidth]{figure/italia.png}
	\label{fig:population}
}
\hfill
\subfloat[Tuscany and the municipality of \emph{Bagni di Lucca} (in yellow)]{
	\includegraphics[width = 0.3 \linewidth]{figure/toscana.png}
	\label{fig:tourismflows}
}
\hfill
\subfloat[Location of the \emph{Casoli}  village in the \emph{Bagni di Lucca} municipality, bordered in yellow. The small town of \emph{Bagni di Lucca} is visible located a few kilometers west of \emph{Casoli} ]{
	\includegraphics[width = 0.3 \linewidth]{figure/bagni_di_lucca.png}
	\label{fig:tourismflows}
}
\caption[The region of \emph{Casoli}]{The region of \emph{Casoli}}
\label{fig:toscana-casoli}
\end{figure}

The latter is also a SPA (Special Protection Area) and the \textit{Orrido di Botri} State Reserve is located within it \cite{nat00}. It is therefore a natural heritage with a fragile balance that needs to be preserved.

@@

In the last few years the community opened two entirely new trails:
\begin{itemize}
	\item the \textit{Alta Via dei Pastori} (2019), a ring around \textit{Monte Prato Fiorito} that takes up the ancient grazing route, and
	\item the \textit{Sentiero degli Avi} (2020), a ring that from \textit{Montefegatesi}
	reaches \textit{Monte Coronato} \cite{pin21}
\end{itemize}

Another recently developed project (2019) is the expansion of the Saint Bartholomew's Path, which runs through the territory of \textit{Pistoia}, with a
variant that from \textit{Popiglio} continues in five stages in the municipality of \emph{Bagni di Lucca} to \textit{Pieve di Controne}, acting as the '\emph{Lucca} gateway' to the Path \cite{camsb}.

@@

The way to improve the valorization and enjoyment of this area applies to a slow tourism approach. This approach can involve the community, especially in the southern part of the municipality, which is still less frequented, more hilly, and therefore less traveled by the network of trails. We propose the creation of geo-itineraries characterized by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.

@@

Ceramic fragments dated between the 3rd century B.C. and the 1st-2nd century A.D. prove that Ligurian populations occupied the region scattered and in small nuclei. The dedication of the Latin colony of \textit{Lucca} in 180 B.C. marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.

@@

Table \ref{tab:tourism} provides statistical data concerning tourist flows and accommodation in \emph{Bagni di Lucca}. Notably, the tourist density of 106.41 tourists per $km^2$ is higher than the population density (33.95 inhabitants per $km^2$).

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}}
% MDPI internal command: Title for citation in the left column
\TitleCitation{A sustainable approach to tourist signage on heritage trails}
% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0009-0003-1136-570X} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0002-9734-2044} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorC}{0000-0002-5542-7891} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{ Maria Grazia Deri $^{2,}$*\orcidC{}, Letizia Chiti $^{3}$\orcidA{}, Augusto Ciuffoletti $^{1}$\orcidB{} }

@@

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Maria Grazia Deri, Letizia Chiti, and Augusto Ciuffoletti}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{ Deri, M.G.; Chiti, L.; Ciuffoletti, A.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad University of Pisa (retired); augusto.ciuffoletti@unipi.it\\
$^{2}$ \quad University of Pisa; mariagrazia.deri@sp.unipi.it\\
$^{3}$ \quad University of Pisa; l.chiti2@studenti.unipi.it}

% Contact information of the corresponding author
\corres{Correspondence: augusto.ciuffoletti@unipi.it}

% Current address and/or shared authorship
%\firstnote{Current address: Affiliation 3.} 
%\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Guiding the visitor to appreciate the sights during an excursion is crucial to building a memorable experience. In the literature, we find many articles exploring the use of sophisticated technologies towards such a goal. The case of tourism proposals for inland areas exhibits a noteworthy aspect, since the presence of human artifacts and the signage may damage the experience, and introduce pollutants as well. This paper analyzes the signage for a natural trail putting into practice a holistic methodology that takes into account all aspects of the problem, starting from the touristic vocation of the area and the social fabric. The analysis indicates slow, community-involved tourism as the preferable target, with extreme requirements of sustainability in the social, economic, and environmental domains for the signage solution. Appropriate governance guidelines are given to implement the proposal. Footprint evaluations address the QR-code technology, which is used to design a real-scale solution finally deployed on the field.}
%\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results that are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

% Keywords
\keyword{Inland areas; heritage trail; digital sustainability; slow tourism; destination governance; community-involved tourism; cultural heritage; holistic approach; QR-code; trail signage}

@@

%\tableofcontents

@@

Although an official decision is still pending, there is a widely agreed proposal for the naming of the time that we are living as the Anthropocene.  Behind this proposal is the recognition that humans have the potential to significantly alter on a global scale the environment in which they live.

The problem with the above fact is that we humans are not able to understand the long-term consequences of deploying such a potential, nor even to keep it under control. Even worse, such capacities are often used with the aim of short-term local effects, ignoring long-term global ones, with the risk of long-term global deterioration. This attitude pervades all activities, starting with the most basic ones such as the supply of food and homes, and is often influenced by economic profit.

@@

To manage this situation, international agencies promote the keyword "sustainability" as a guideline for the conservation of our habitat. Jeffry Ramsey et al. \cite{ram15a} consider the concept behind this word as "vague and disputed but not meaningless" and that cannot be used without a concrete framework, especially in a normative context. To solve the problem of measuring sustainability, Tom Kuhlman et al. \cite{kuh10a} point out an inherent conceptual tension between growth and stability. This tension has been attenuated by a change in the meaning of the word, which has gone from focusing on preservation for an indefinite future to a concept that includes three "pillars", society, economy, and environment, as determinants that need to be protected in the present to promote sustainable development in the future. The authors relate artificial and environmental goods, discriminating the two cases where a loss in the latter can or cannot be compensated by an increase in the former, introducing the terms of {\em weak} and {\em strong} sustainability.

The promotion of cultural heritage is situated in such a complex scenario. As noted by Jorge Otero in \cite{ote22a}, local communities benefit from the promotion of heritage: conservation activity generates social growth, while tourism provides an extra income that can be reinvested in promotion. However, just as in the case of ski resorts, there is a risk of compromising the heritage itself. This interaction raises a question of sustainability also for heritage valorization initiatives, two issues that, instead,  are often considered in a virtual synergy. As stated in the document titled 'World Heritage and Sustainable Development' by UNESCO \cite{lab17a}, calling for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".

This research stems from such a tough contradiction in trying to answer the question: is there a way to unlock the potential of cultural heritage while at the same time limiting the damage to a social, economic, and environmental equilibrium?

The question arises in the course of the Underlandscape project \citeWeb{underlandscape} funded by the Italian Ministry of Research aimed at applying cutting-edge technologies to the investigation of karst caves that may have been inhabited, from prehistory to today. Among the expected results of the project is the definition of a strategy to disseminate the research outcomes allowing non-specialists to visit and enjoy the results, thus contributing to the economy of inland areas that are severely exposed to de-population. A sustainable strategy is even more crucial as the geological environment of interest is particularly fragile, as noted by Aleksander Antic\`c et al. in \cite{ant20a}.

This document provides a conceptual solution for a specific aspect: heritage trail signage. The solution is conceptual in that it tries to follow a guideline that takes into account many aspects and provides a practical way that is adaptable to other environments.

@@

The functions of signage are two:
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map. The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream". A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}). The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
\item a sign should explain the reasons for interest in a site. This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}). In this sense, the CAI signal above is not sufficient. A simple board with a site name fits only when the site is sufficiently popular. Otherwise, a more structured message should be used to explain its relevance.
\end{itemize}

In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}). We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.

The purpose of such an installation is guiding the {\em user} on a tour that includes urban streets, buildings, natural trails, and caves (this latter being the topic of the project giving financial support to our research). The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as profitable and enjoyable as possible.

A traditional approach makes use of physical information boards with graphical or textual content, as in figure \ref{fig:traditional}. Our study starts by pointing out the issues related to such a solution:

\begin{itemize}
	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance;
	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports;
	\item environmental impact: to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site; 
	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons;
	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well;
	\item update limits: to update the content, the plaque must be replaced;
	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation.
\end{itemize}

Such points motivate an investigation to find alternative ways.

@@

We characterize our approach as an instance of {\em weak sustainability}: we do not preclude intervention in the environment, but the impact of such intervention must be better than that related to traditional signs. To this end, we include in our solution tools that are not part of the site but remain with the user.

During the last decades, we witnessed the diffusion of smartphone devices, and we have no reason to expect a change in such a trend. Smartphones enhance the communication capabilities of individuals by allowing them to receive sound, visual, and tactile interactions. The relevance of such capabilities for the improvement of a tourist experience has been widely investigated, either in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a} to find ways to exploit such tools. As we did with physical information boards, we highlight the limitations of this technology, particularly when it comes to outdoor activities.

@@

Regarding \textit{usability}, the basic recommendation is to keep the operation within the experience of the majority of users, without requiring them to familiarize themselves with new applications.

@@

On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the term introduced by Pearce and Gretzel in \cite{pea12a}, may evoke contrasting feelings, from rewarding to threatening.

@@

The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support, taking into account its sustainability "by design". Once a range of relevant solutions is identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies the feasibility of a specific solution.

We emphasize that this article is not meant to quantify user satisfaction or the economic revenue associated with the specific solution. Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve a thorough investigation. We aim at isolating an issue, proposing a sustainable strategy for its solution, and implementing a proof of concept for it.

@@

%We aim to design sustainable and effective smartphone-based signage.

The wireless capabilities of a smartphone are the focus of a straightforward approach. Given the premise that the Internet is assumed to be unreachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest. Small servers connected to the network would host specific Web services. The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single-chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$. It can provide a WiFi Access Point as well as Web content. The ESP8266 has a capacity of 32KB, which is sufficient for explanatory text and a low-resolution image. Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.

Such a solution is severely limited by the power supply dependency. A radio transmitter is a rather power-consuming device, in the range of Watts. Even if intermittent, the operation of a battery-operated networking device cannot be guaranteed for long periods and the host organization should consider power harvesting, which negatively contributes to the economic cost and environmental footprint of the device.

@@

An alternative consists of the utilization of passive devices, like Near Field Communication (NFC) transmitters. The transmitting device is flat, the size of a coin, and costs less than one dollar per piece. To receive the content the smartphone must be very close to the NFC tag. The power needed to operate the radio is drained from the smartphone so that the transmitter does not depend on batteries. The NFC device capacity is in the range of the KBytes, nearly a page on this journal. Engraving content onto an NFC tag is a straightforward task that requires a smartphone and an ad-hoc app.

The NFC technology is currently very diffused, although reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text. The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues. Such capabilities do not need an Internet connection.

@@

\begin{itemize}
	\item {\em Equitable use} is closely related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone,
	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety. The deliberate or accidental release of the passive device into the environment determines a minor pollution,
	\item{\em Low physical effort} holds, although the user needs to carry a smartphone,
	\item{\em Size and Space for approach and use} need to be carefully considered. In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort.
\end{itemize}

@@

Our study aims to demonstrate a sustainable solution in a concrete setting through a proof of concept. In order to follow a holistic approach, we must start with the definition of the operation framework.

This section is dedicated to describing the economic and social context, as well as the historical background that characterizes the heritage resource we want to promote.

The geographical area of interest is the surroundings of \emph{Casoli}, a small village in a mountainous region in the north of Tuscany, Italy. The area falls within the municipality of \emph{Bagni di Lucca}, in the province of \emph{Lucca}. In 2021, the archaeological team of our project realized a thematic map assessing the reasons of interest for the heritage and the geomorphological features of the area.

@@

\emph{Bagni di Lucca} is located on the north-eastern boundary of the Province of \emph{Lucca}, in \emph{Val di Lima}, and is part of the \textit{Media Valle del Serchio} district. With its mountains it marks the historical border with the \textit{Modena} and \textit{Pistoia} area, it is very rich in potential for its naturalistic, archaeological, and historical heritage, both expressed and as yet unexpressed. Thanks to the archaeological map the main sites of interest from prehistoric to contemporary times in this municipality are cataloged, photographed, and georeferenced.

From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress the area with a peculiar geomorphology impervious despite the modest hilly elevations.

The \emph{Lima} stream has characterized the history of \emph{Bagni di Lucca} since ever: the valley's ancient river terraces contain some of the earliest evidence of human presence, such as caves and rock shelters frequented since the Palaeolithic age; the manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited since the Middle Age until the 1980s \cite{men76, bed05, ser21}.

On the naturalistic side, the region of \emph{Bagni di Lucca} encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \citeWeb{nat00} European Economic Community (EEC) initiative.

There are three SCIs-SACs (Sites of Community Interest and Special Areas of Conservation) located in the area north of the \emph{Lima} stream, corresponding to the Apennine portion, covering 23\% of the municipal surface:
\begin {itemize}
\item the limestone areas of \emph{Val di Lima} and \emph{Balzo Nero}; 
\item \emph{Monte Prato Fiorito}-\emph{Monte Coronato}-\emph{Valle dello
Scesta}; 
\item \emph{Orrido di Botri}.
\end{itemize}

@@

\begin{figure}
\subfloat[Italy and Tuscany, bordered in red]{
	\includegraphics[width = 0.3\linewidth]{figure/italia.png}
}
\hfill
\subfloat[Tuscany and the municipality of \emph{Bagni di Lucca} (in yellow)]{
	\includegraphics[width = 0.3 \linewidth]{figure/toscana.png}
}
\hfill
\subfloat[Location of the \emph{Casoli}  village in the \emph{Bagni di Lucca} municipality, bordered in yellow. The small town of \emph{Bagni di Lucca} is visible located a few kilometers west of \emph{Casoli} ]{
	\includegraphics[width = 0.3 \linewidth]{figure/bagni_di_lucca.png}
}
\caption[The region of \emph{Casoli}]{The region of \emph{Casoli}}
\label{fig:toscana-casoli}
\end{figure}

The latter is also a SPA (Special Protection Area) and the \textit{Orrido di Botri} State Reserve is located within it \citeWeb{nat00}. It is therefore a natural heritage with a fragile balance that needs to be preserved.

@@

In the last few years, the community opened two entirely new trails:
\begin{itemize}
	\item the \textit{Alta Via dei Pastori} (2019), a ring around \textit{Monte Prato Fiorito} that takes up the ancient grazing route, and
	\item the \textit{Sentiero degli Avi} (2020), a ring that from \textit{Montefegatesi}
	reaches \textit{Monte Coronato} \cite{pin21}
\end{itemize}

Another recently developed project (2019) is the expansion of the Saint Bartholomew's Path, which runs through the territory of \textit{Pistoia}, with a
variant that from \textit{Popiglio} continues in five stages in the municipality of \emph{Bagni di Lucca} to \textit{Pieve di Controne}, acting as the '\emph{Lucca} gateway' to the Path \citeWeb{camsb}.

@@

The way to improve the valorization and enjoyment of this area applies to a slow tourism approach. This approach can involve the community, especially in the southern part of the municipality, which is still less frequented, more hilly, and therefore less traveled by the network of trails. We envision the creation of geo-itineraries characterized by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.

@@

Ceramic fragments dated between the 3rd century BC and the 1st-2nd century AD prove that Ligurian populations occupied the region scattered and in small nuclei. The dedication of the Latin colony of \textit{Lucca} in 180 BC marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.

@@

Table \ref{tab:tourism} provides statistical data concerning tourist flows and accommodation in \emph{Bagni di Lucca}. Notably, the tourist density of 106.41 tourists per $km^2$ is higher than the population density (33.95 inhabitants per $km^2$). This information should be considered in local planning to enhance the number of residents.
